---
title: "Preprossessing data analysis"
output: html_document
date: "2026-03-25"
---


# Goals
1. Remove chloroplasts
2. Remove mitochondria
3. Remove ASVs from negative controls, and remove negative control samples
4. Evaluate sequencing occurancy with the mock community (positive control)
5. 

# Output: 
1. Cleaned up phyloseq object


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      # send any figure output to this folder 
                      fig.path = "../figures/02_PreProcessing/")
# Set the seed for reproducibility
set.seed(238428)

# What time did we start running this script? 
start_time <- Sys.time()

# Load packages 
pacman::p_load(devtools, phyloseq, Biostrings, tidyverse, dada2, install = FALSE)

# Load the data 
load("/local/workdir/yz3272/Biodiversity_Salinity_16S/data/02_AssignASVs/raw_physeq.RData")

# Look at the data 
raw_physeq

```

```{r Clean up the data}
# Let's take a quick look at the tax table 
# View(tax_df) # Let's see where chloroplasts are and also mitochondria

# How to access what is in a phyloseq table
view(tax_table(raw_physeq))
view(otu_table(raw_physeq))

# Make new physeq without chloroplasts
noChloros_physeq <- 
  raw_physeq %>% 
  # rm chloroplasts, which are Order
  subset_taxa(Order != "Chloroplast" | is.na(Order))
  
# How many taxa were chloroplasts? 
num_chloro_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
num_chloro_ASVs #248 chloroplast ASVs

# Intuition chek 
#noChloros_physeq %>%
#  tax_table() %>%
#  data.frame() %>%
#  View()

# remove mitochondria 
noChlorosMitos_physeq <- 
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family))

# How many mitochondrial ASVs? 
num_mito_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)
num_mito_ASVs ##74 mito ASVs

# Let's take a quick look 
noChlorosMitos_physeq

# How many total asvs were removed from chloros and mitos? 
ntaxa(raw_physeq) - ntaxa(noChlorosMitos_physeq)

# proportion of asvs kept? 
ntaxa(noChlorosMitos_physeq)/ntaxa(raw_physeq)
```
```{r Negative controls}
# create a vector of samples that were negative controls 
control_samples <- c("WaterControl", "022um-Control", "3um-Control", "DNA-Ext-Control")

# Make a new phyloseq object of only the control samples 
control_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(., names %in% control_samples) %>%
  # force remove ASVs with 0 counts 
  prune_taxa(taxa_sums(.) > 0, .)

# Let's take a look
control_physeq

# vector of control asvs
control_ASVs <- 
  control_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV) %>%
  as.vector()

# Visualize raw abundance values of control ASVs
control_physeq %>%
  plot_bar(., "ASV", fill = "ASV")

############# Check the control ASVs in the samples 
# Make new phyloseq object with only the control ASVs and ALL samples 
controlASV_physeq <- 
  noChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% control_ASVs$ASV)

# Inspect it 
controlASV_physeq

# Let's take a look at the abundances 
controlASV_physeq %>%
  plot_bar(., "ASV", fill = "ASV", facet_grid = Sample_or_Control~.)

view(tax_table(controlASV_physeq))
#see interpretation below

# Would only like to keep ASV_0019 and ASV_0055, everything else is trash
controlASVs_toREMOVE <- dplyr::setdiff(control_ASVs$ASV, c("ASV_0019", "ASV_0055"))
# double check 
length(controlASVs_toREMOVE)

### Make my new phloseq object 
noControlsChlorosMitos_physeq <- 
  noChlorosMitos_physeq %>%
  # Actually remove the 29 control ASVs 
  subset_taxa(., !(ASV %in% controlASVs_toREMOVE)) %>% #subset to remove the control ASVs to remove
  # Remove control samples 
  subset_samples(., !(names %in% control_samples))

# Inspect no controls object 
noControlsChlorosMitos_physeq

# After filtering, how many ASVs were removed? 
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noControlsChlorosMitos_physeq)
total_ASVs_rm


```
### Interpretation: 
ASV 3 looks like contamination, but ASV 0019 and 0055 are likely real taxa (water microbes) and need to check in the actual data of samples, see who they are, whether they are abundant water taxa, if not, these two and all other empty ASVs can be removed.After checking, 19 and 55 are indeed water microbes, we will keep these two. 

```{r Evaluate mock community}
# Make mock phyloseq object 
mock_physeq <- 
  noControlsChlorosMitos_physeq %>%
  subset_samples(., names == "MockZymoPos") %>%
  prune_taxa(taxa_sums(.) > 0, .)

# Inspect it 
mock_physeq

tax_table(mock_physeq)

# write us a messages summarizing this
cat("DADA2 inferred", ntaxa(mock_physeq), "ASVs present in the mock community.")

####### Load in the mock reference sequence from Zymo 
mock_ref <- getSequences("data/Lecture/03_PreProcessing_Lecture/mock_amplicons.fasta")
names(mock_ref)

### Pull out seqs from Sequenced Mock community
mock_seqs <- 
  mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(ASV, ASVseqs) 

# Inspect
head(mock_seqs)

# Which ASVs match the reference? 
matched_asvs <- 
  mock_seqs %>%
  rowwise() %>%
  # check each ASV if they match, TRUE, if not = FALSE
  mutate(Match = any(grepl(ASVseqs, mock_ref))) %>%
  # Create a vector of names that matched 
  pull(ASV)


# Evaluate which ones do NOT match?? 
# What's their abundances? 
# This help provide error rate 
cat(length(matched_asvs), "ASVs were exact matches to the expected reference sequence.")

# How many sequences were in the mock? 
mock_SeqDepth <- 
  mock_physeq %>%
  otu_table() %>%
  data.frame() %>%
  colSums()


# Who is in the sequenced mock community? 
mock_physeq %>%
  tax_table() %>%
  data.frame() %>%
  dplyr::select(Genus, Species)

names(mock_ref)

# Curious: What are the mock counts in the actual samples 
noControlsChlorosMitos_physeq %>%
  subset_taxa(., ASV %in% matched_asvs) %>%
  otu_table() %>%
  t()
```

```{r}

```



